---
title: "Drainage Area Estimation"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drainage Area Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(nhdplusTools)

local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")
if(local) {
  cache_path <- file.path(nhdplusTools_data_dir(), "da_v_cache")
} else {
  cache_path <- tempdir()
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = local,
  cache = local,
  cache.path = (cache_path)
)

oldoption <- options(scipen = 9999)
```

`get_drainage_area_estimates()` combines HUC12 areas upstream of HUC
outlets with NHDPlusV2 catchment areas for the gap between the outlet
and the nearest HUC12 boundaries. Non-contributing areas captured in
HUC12 boundaries are included, making these estimates more
representative than purely network-derived totals.

This vignette runs the function on five basins that span a range of
hydrologic settings — from a well-determined humid basin to arid
systems, glacial prairie, and an endorheic (closed) basin.

## Fetch and Cache Results

Each basin result is stored in a separate RDS file in the nhdplusTools
data directory so that subsequent runs skip the web service calls.

```{r fetch_data, message = TRUE}
library(sf)

data_dir <- nhdplusTools_data_dir()
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

# Define the five study sites
sites <- list(
  black_earth = list(featureSource = "nwissite",
    featureID = "USGS-05406500"),
  french_broad = list(featureSource = "nwissite",
    featureID = "USGS-03451500"),
  brazos = list(featureSource = "nwissite",
    featureID = "USGS-08116650"),
  james = list(featureSource = "nwissite",
    featureID = "USGS-06468000"),
  malheur = st_sfc(st_point(c(-118.8, 43.3)), crs = 4326)
)

fetch_or_load <- function(name, start) {
  rds_path <- file.path(data_dir,
    paste0("da_est_", name, ".rds"))
  if(file.exists(rds_path)) {
    message("Loading cached: ", name)
    return(readRDS(rds_path))
  }
  message("Fetching: ", name)
  result <- tryCatch(
    get_drainage_area_estimates(start, catchments = TRUE),
    error = function(e) {
      warning("Failed for ", name, ": ", conditionMessage(e),
        call. = FALSE)
      NULL
    })
  if(!is.null(result)) saveRDS(result, rds_path)
  result
}

da_results <- Map(fetch_or_load, names(sites), sites)
da_results <- Filter(Negate(is.null), da_results)
```

## Return Structure

Each result is a list with scalar drainage area estimates and spatial
data frames. Here are the elements for Black Earth Creek:

```{r structure}
names(da_results$black_earth)
```

The scalar estimates (square kilometers) across all five basins:

```{r summary_table}
basin_labels <- c(
  black_earth = "Black Earth Creek",
  french_broad = "French Broad",
  brazos = "Brazos at Rosharon",
  james = "James River",
  malheur = "Malheur Lake"
)

summary_df <- data.frame(
  basin = basin_labels[names(da_results)],
  network_da = vapply(da_results, \(x) x$network_da_sqkm, numeric(1)),
  da_huc12 = vapply(da_results, \(x) x$da_huc12_sqkm, numeric(1)),
  contrib_huc12 = vapply(da_results,
    \(x) x$contrib_da_huc12_sqkm, numeric(1)),
  da_huc10 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc10_sqkm), NA_real_, x$da_huc10_sqkm),
    numeric(1)),
  da_huc08 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc08_sqkm), NA_real_, x$da_huc08_sqkm),
    numeric(1)),
  nhdplushr = vapply(da_results,
    \(x) ifelse(is.na(x$nhdplushr_network_dasqkm), NA_real_,
      x$nhdplushr_network_dasqkm),
    numeric(1))
)

knitr::kable(summary_df, digits = 1,
  col.names = c("Basin", "Network DA", "HUC12 DA",
    "Contributing DA", "HUC10 DA", "HUC08 DA", "NHDPlusHR DA"),
  caption = "Drainage area estimates (sq km) by basin and method")
```

Spatial elements in each result:

```{r spatial_elements}
be <- da_results$black_earth

cat("start_feature:", class(be$start_feature)[1], "-",
  nrow(be$start_feature), "rows\n")
cat("hu12_by_huc12:", class(be$hu12_by_huc12)[1], "-",
  nrow(be$hu12_by_huc12), "rows\n")
cat("hu12_by_huc10:",
  if(is.null(be$hu12_by_huc10)) "NULL"
  else paste(class(be$hu12_by_huc10)[1], "-",
    nrow(be$hu12_by_huc10), "rows"), "\n")
cat("hu12_by_huc08:",
  if(is.null(be$hu12_by_huc08)) "NULL"
  else paste(class(be$hu12_by_huc08)[1], "-",
    nrow(be$hu12_by_huc08), "rows"), "\n")
cat("extra_catchments:", class(be$extra_catchments)[1], "-",
  nrow(be$extra_catchments), "rows\n")
cat("split_catchment:", class(be$split_catchment)[1], "-",
  nrow(be$split_catchment), "rows\n")
cat("all_network:", class(be$all_network)[1], "-",
  nrow(be$all_network), "rows\n")
cat("all_catchments:", class(be$all_catchments)[1], "-",
  nrow(be$all_catchments), "rows\n")
cat("hu12_outlet:", class(be$hu12_outlet)[1], "-",
  nrow(be$hu12_outlet), "rows\n")
```

## Black Earth Creek — HUC12 Non-Contributing Area

```{r plot_black_earth, fig.width = 7, fig.height = 6}
library(ggplot2)

be <- da_results$black_earth

hu12 <- be$hu12_by_huc12
hu12$pct_noncontrib <- hu12$ncontrb_sqkm / hu12$dasqkm * 100

start_pt <- st_geometry(be$start_feature)
if(!inherits(start_pt, "sfc_POINT"))
  start_pt <- st_centroid(start_pt)
start_pt <- st_transform(st_sf(geometry = start_pt), st_crs(hu12))

ggplot() +
  geom_sf(data = hu12, aes(fill = pct_noncontrib),
    color = "gray40", linewidth = 0.3) +
  scale_fill_gradient(low = "#FFFDD0", high = "#8B4513",
    name = "Non-contributing\narea (%)") +
  geom_sf(data = be$extra_catchments, fill = "lightblue",
    color = "steelblue", alpha = 0.5, linewidth = 0.3) +
  geom_sf(data = be$split_catchment, fill = NA,
    color = "red", linewidth = 0.5, linetype = "dashed") +
  geom_sf(data = start_pt, shape = 17, size = 3, color = "black") +
  geom_sf(data = be$hu12_outlet, shape = 4, size = 2,
    color = "darkred") +
  labs(
    title = "Black Earth Creek - Drainage Area Estimation",
    subtitle = paste0(
      "Network DA: ", round(be$network_da_sqkm, 1), " sq km | ",
      "HUC12 DA: ", round(be$da_huc12_sqkm, 1), " sq km")
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r teardown, include = FALSE}
options(oldoption)

if(Sys.getenv("BUILD_VIGNETTES") != "TRUE") {
  unlink(cache_path, recursive = TRUE)
}
```
