---
title: "Drainage Area Estimation"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drainage Area Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo = FALSE}
/* Override html_vignette narrow body to allow wider figures */
body, div.main-container, div.body, div.contents {
  max-width: 1600px !important;
  width: 95% !important;
}
img {
  max-width: 100% !important;
  height: auto !important;
}
.vignette-note {
  background: #f8f8f8;
  border-left: 3px solid #ccc;
  padding: 8px 12px;
  margin: 8px 0 16px 0;
  font-size: 0.9em;
}
```

```{r setup, include = FALSE}
library(nhdplusTools)

build_local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 16,
  fig.height = 9,
  dpi = 110,
  out.width = "100%",
  eval = build_local
)

huc12 <- NULL
f <- "../../Data/nhdp/NHDPlusNationalData/NHDPlusV21_National_Seamless_Flattened_Lower48.gdb/"
if(file.exists(f))
  huc12 <- sf::read_sf(f, "HUC12")

oldoption <- options(scipen = 9999)
```

`get_drainage_area_estimates()` combines HUC12 areas upstream of HUC
outlets with NHDPlusV2 catchment areas for the gap between the outlet
and the nearest HUC12 boundaries. Non-contributing areas captured in
HUC12 boundaries are included, making these estimates more
representative than purely network-derived totals.

This vignette runs the function on five basins that span a range of
hydrologic settings -- from a well-determined humid basin to arid
systems, glacial prairie, and an endorheic (closed) basin.

## Fetch and Cache Results

Each basin result is stored in a separate RDS file in the nhdplusTools
data directory so that subsequent runs skip the web service calls.

```{r fetch_data, message = TRUE}
library(sf)

data_dir <- nhdplusTools_data_dir()
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

# Define the five study sites
sites <- list(
  black_earth = list(featureSource = "nwissite",
    featureID = "USGS-05406500"),
  french_broad = list(featureSource = "nwissite",
    featureID = "USGS-03451500"),
  brazos = list(featureSource = "nwissite",
    featureID = "USGS-08116650"),
  james = list(featureSource = "nwissite",
    featureID = "USGS-06468000"),
  malheur = st_sfc(st_point(c(-118.8, 43.3)), crs = 4326)
)

fetch_or_load <- function(name, start) {
  rds_path <- file.path(data_dir,
    paste0("da_est_", name, ".rds"))
  if(file.exists(rds_path)) {
    message("Loading cached: ", name)
    return(readRDS(rds_path))
  }
  message("Fetching: ", name)
  result <- tryCatch(
    get_drainage_area_estimates(start, catchments = TRUE,
      huc12_data = huc12),
    error = function(e) {
      warning("Failed for ", name, ": ", conditionMessage(e),
        call. = FALSE)
      NULL
    })
  if(!is.null(result)) saveRDS(result, rds_path)
  result
}

da_results <- Map(fetch_or_load, names(sites), sites)
da_results <- Filter(Negate(is.null), da_results)
```

```{r fetch_flowlines, message = TRUE}
fetch_flowlines <- function(name, da_result) {
  rds_path <- file.path(data_dir, paste0("fl_", name, ".rds"))
  if(file.exists(rds_path)) {
    message("Loading cached flowlines: ", name)
    return(readRDS(rds_path))
  }
  message("Fetching flowlines: ", name)
  comids <- da_result$all_network$comid
  fl <- tryCatch(
    get_nhdplus(comid = comids, realization = "flowline"),
    error = function(e) {
      warning("Flowline fetch failed for ", name, ": ",
        conditionMessage(e), call. = FALSE)
      NULL
    })
  if(!is.null(fl)) saveRDS(fl, rds_path)
  fl
}

flowlines <- Map(fetch_flowlines, names(da_results), da_results)
flowlines <- Filter(Negate(is.null), flowlines)
```

## Return Structure

Each result is a list with scalar drainage area estimates and spatial
data frames. Here are the elements for Black Earth Creek:

```{r structure}
names(da_results$black_earth)
```

The scalar estimates (square kilometers) across all basins:

```{r summary_table}
basin_labels <- c(
  black_earth = "Black Earth Creek",
  french_broad = "French Broad",
  brazos = "Brazos at Rosharon",
  james = "James River",
  malheur = "Malheur Lake"
)

summary_df <- data.frame(
  basin = basin_labels[names(da_results)],
  network_da = vapply(da_results, \(x) x$network_da_sqkm, numeric(1)),
  da_huc12 = vapply(da_results, \(x) x$da_huc12_sqkm, numeric(1)),
  contrib_huc12 = vapply(da_results,
    \(x) x$contrib_da_huc12_sqkm, numeric(1)),
  da_huc10 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc10_sqkm), NA_real_, x$da_huc10_sqkm),
    numeric(1)),
  da_huc08 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc08_sqkm), NA_real_, x$da_huc08_sqkm),
    numeric(1)),
  nhdplushr = vapply(da_results,
    \(x) ifelse(is.na(x$nhdplushr_network_dasqkm), NA_real_,
      x$nhdplushr_network_dasqkm),
    numeric(1))
)

knitr::kable(summary_df, digits = 1,
  col.names = c("Basin", "Network DA", "HUC12 DA",
    "Contributing DA", "HUC10 DA", "HUC08 DA", "NHDPlusHR DA"),
  caption = "Drainage area estimates (sq km) by basin and method")
```

```{r helpers, include = FALSE}
library(ggplot2)

# USGS USTopo provider for maptiles. The "?f=jpg" suffix is a no-op
# query parameter that gives maptiles a recognized extension hint.
usgs_topo_provider <- list(
  src = "USGS_USTopo",
  q = paste0("https://basemap.nationalmap.gov/arcgis/rest/services/",
    "USGSTopo/MapServer/tile/{z}/{y}/{x}?f=jpg"),
  sub = NA,
  cit = "USGS The National Map: USGSTopo"
)

# Cache directory for tiles
tile_cache <- file.path(nhdplusTools_data_dir(), "da_v_tiles")
dir.create(tile_cache, recursive = TRUE, showWarnings = FALSE)

# Compute a target tile zoom level so the rendered tiles are sharp
# at the figure size we use. Web Mercator world width is 40075016 m
# and tile size is 256 px. The figure pixel width drives the
# resolution we want from the tile mosaic.
target_zoom <- function(sf_obj, fig_px = 1700) {
  bb <- st_bbox(sf_obj)
  span_m <- max(bb["xmax"] - bb["xmin"], bb["ymax"] - bb["ymin"])
  if(!is.finite(span_m) || span_m <= 0) return(10L)
  z <- log2(40075016 * fig_px / (256 * span_m))
  as.integer(min(max(round(z), 4L), 15L))
}

# Fetch USGS topo tiles for an sf bbox. Returns a list with a color
# matrix (suitable for annotation_raster) and the tile bounding box.
# Uses annotation_raster (not geom_raster) so the fill aesthetic
# stays free for choropleth use.
fetch_topo_raster <- function(sf_obj, zoom = NULL) {
  bb <- st_bbox(sf_obj)
  pad_x <- (bb["xmax"] - bb["xmin"]) * 0.05
  pad_y <- (bb["ymax"] - bb["ymin"]) * 0.05
  bb["xmin"] <- bb["xmin"] - pad_x
  bb["xmax"] <- bb["xmax"] + pad_x
  bb["ymin"] <- bb["ymin"] - pad_y
  bb["ymax"] <- bb["ymax"] + pad_y
  bb_sf <- st_as_sfc(bb, crs = st_crs(sf_obj))
  if(is.null(zoom))
    zoom <- target_zoom(st_transform(bb_sf, 3857))
  args <- list(
    x = bb_sf, provider = usgs_topo_provider, crop = TRUE,
    cachedir = tile_cache, zoom = zoom
  )
  tiles <- tryCatch(
    do.call(maptiles::get_tiles, args),
    error = function(e) {
      warning("Tile fetch failed: ", conditionMessage(e),
        call. = FALSE)
      NULL
    })
  if(is.null(tiles)) return(NULL)
  rgb_arr <- terra::as.array(tiles)
  if(length(dim(rgb_arr)) != 3 || dim(rgb_arr)[3] < 3) return(NULL)
  col_mat <- matrix(
    rgb(rgb_arr[, , 1], rgb_arr[, , 2], rgb_arr[, , 3],
      maxColorValue = 255),
    nrow = dim(rgb_arr)[1], ncol = dim(rgb_arr)[2]
  )
  ext <- terra::ext(tiles)
  list(
    raster = col_mat,
    xmin = ext[1], xmax = ext[2],
    ymin = ext[3], ymax = ext[4]
  )
}

# Add basemap as annotation_raster (does not consume fill scale)
add_topo <- function(p, sf_obj) {
  topo <- fetch_topo_raster(sf_obj)
  if(is.null(topo)) return(p)
  p + annotation_raster(
    topo$raster,
    xmin = topo$xmin, xmax = topo$xmax,
    ymin = topo$ymin, ymax = topo$ymax
  )
}

# Resolve gage point in target CRS
get_gage_pt <- function(da, target_crs) {
  pt <- st_geometry(da$start_feature)
  if(!inherits(pt, "sfc_POINT"))
    pt <- st_centroid(pt)
  st_transform(st_sf(geometry = pt), target_crs)
}

# Pick the spatially-largest of the available HUC12 sets returned by
# get_drainage_area_estimates: hu12_by_huc08 (if present), otherwise
# hu12_by_huc10, otherwise hu12_by_huc12. Largest is determined by
# the sum of HUC12 dasqkm in each set.
largest_hu12_set <- function(da) {
  candidates <- list(
    huc12 = da$hu12_by_huc12,
    huc10 = da$hu12_by_huc10,
    huc08 = da$hu12_by_huc08
  )
  candidates <- Filter(
    function(x) !is.null(x) && nrow(x) > 0, candidates)
  areas <- vapply(candidates,
    function(x) sum(x$dasqkm, na.rm = TRUE), numeric(1))
  candidates[[which.max(areas)]]
}

# Compute a single merged basin polygon for a given HUC level set
# (hu12_by_huc12, hu12_by_huc10, or hu12_by_huc08), unioned with
# extra_catchments and the local catchments around the gage.
# Uses hydroloom::dissolve_polygons() with gap_tolerance to absorb
# sliver gaps between polygons from different services (WBD HUC12s,
# NHDPlusV2 catchments, split-catchment service).
compute_basin_polygon <- function(da, hu_layer, target_crs,
  tol_m = 250) {
  common_crs <- st_crs(hu_layer)
  pieces <- list()
  pieces$hu <- st_geometry(hu_layer)

  if(!is.null(da$extra_catchments) && nrow(da$extra_catchments) > 0)
    pieces$extra <- st_geometry(
      st_transform(da$extra_catchments, common_crs))

  if(!is.null(da$split_catchment) && nrow(da$split_catchment) > 0) {
    sc <- da$split_catchment
    sc_full <- sc[sc$id == "catchment", ]
    if(nrow(sc_full) > 0)
      pieces$catch <- st_geometry(st_transform(sc_full, common_crs))
  }

  all_polys <- st_sf(
    geometry = do.call(c, pieces)
  ) |> st_set_crs(common_crs)

  hydroloom::dissolve_polygons(all_polys,
    gap_tolerance = tol_m,
    max_hole_area = Inf,
    single_polygon = TRUE) |>
    st_transform(target_crs)
}

# Common theme: no axis ticks, labels, or grid
map_theme <- function() {
  theme_void() +
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      plot.title = element_text(hjust = 0, size = 13,
        margin = margin(b = 6))
    )
}

# Add the gray basin underlay to a plot. The polygon represents the
# full estimated drainage area envelope (largest available HUC level
# plus extra catchments and local catchments).
add_basin_underlay <- function(p, basin_poly) {
  p + geom_sf(data = basin_poly,
    fill = "gray30", color = NA, alpha = 0.15,
    inherit.aes = FALSE)
}

# Mode 1: Drainage Area Boundaries
plot_boundaries <- function(da, basin_label) {
  # Reproject working layers to Web Mercator (3857) so the basemap
  # tiles align cleanly.
  target_crs <- st_crs(3857)
  gage_pt <- get_gage_pt(da, target_crs)
  basin_poly <- compute_basin_polygon(da, largest_hu12_set(da), target_crs)

  # Dissolve boundaries at each HUC level. Each HUC level is combined
  # with extra_catchments and split_catchment via compute_basin_polygon
  # so every boundary reaches the gage instead of stopping at the
  # lowest complete HUC outlet.
  boundaries <- list()
  boundaries[["HUC12 (NLDI)"]] <- st_geometry(
    compute_basin_polygon(da, da$hu12_by_huc12, target_crs))
  if(!is.null(da$hu12_by_huc10))
    boundaries[["HUC10"]] <- st_geometry(
      compute_basin_polygon(da, da$hu12_by_huc10, target_crs))
  if(!is.null(da$hu12_by_huc08))
    boundaries[["HUC08"]] <- st_geometry(
      compute_basin_polygon(da, da$hu12_by_huc08, target_crs))
  if(!is.null(da$all_catchments))
    boundaries[["Network catchments"]] <- st_union(
      st_transform(da$all_catchments, target_crs))

  ltypes <- c("solid", "dashed", "dotted", "twodash")

  bnd_sf <- do.call(rbind, lapply(seq_along(boundaries), function(i) {
    st_sf(source = names(boundaries)[i],
      geometry = boundaries[[i]])
  }))

  hu12_pp <- if(!is.null(da$hu12_outlet))
    st_transform(da$hu12_outlet, target_crs) else NULL

  # Combine point layers into one sf so a shape scale can label both
  pts <- st_sf(
    label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt)
  )
  if(!is.null(hu12_pp)) {
    pts <- rbind(
      pts,
      st_sf(label = "HUC12 outlet",
        geometry = st_geometry(hu12_pp))
    )
  }

  p <- ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = bnd_sf, aes(linetype = source),
      fill = NA, color = "gray10", linewidth = 0.8,
      inherit.aes = FALSE) +
    scale_linetype_manual(
      values = setNames(ltypes[seq_along(boundaries)],
        names(boundaries)),
      name = "DA boundary") +
    geom_sf(data = pts, aes(shape = label, size = label,
      color = label),
      fill = "white", stroke = 1.4, inherit.aes = FALSE) +
    scale_shape_manual(
      values = c("Stream gage / outlet" = 17, "HUC12 outlet" = 4),
      name = NULL) +
    scale_size_manual(
      values = c("Stream gage / outlet" = 5, "HUC12 outlet" = 3),
      name = NULL) +
    scale_color_manual(
      values = c("Stream gage / outlet" = "black",
        "HUC12 outlet" = "darkred"),
      name = NULL) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label, "-- Drainage Area Boundaries")) +
    map_theme()

  p
}

# Mode 2: Stream Network
plot_network <- function(da, fl, basin_label) {
  target_crs <- st_crs(3857)
  gage_pt <- get_gage_pt(da, target_crs)
  fl <- st_transform(fl, target_crs)
  basin_poly <- compute_basin_polygon(da, largest_hu12_set(da), target_crs)

  # fcode 46003 = intermittent, 46007 = ephemeral, 46006 = perennial
  fl$flow_class <- ifelse(
    fl$fcode %in% c(46003L, 46007L), "Intermittent/Ephemeral",
    "Perennial/Other"
  )
  fl$lwd <- pmin(pmax(fl$streamorde, 1) / 3, 1.5)

  pts <- st_sf(label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt))

  ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = fl,
      aes(color = flow_class, linewidth = lwd),
      show.legend = "line", inherit.aes = FALSE) +
    scale_color_manual(
      values = c("Perennial/Other" = "steelblue",
        "Intermittent/Ephemeral" = "lightskyblue"),
      name = "Flow class") +
    scale_linewidth_identity() +
    ggnewscale_color(pts, target_crs) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label, "-- Stream Network")) +
    map_theme()
}

# Helper to add the gage point as a labeled layer using a shape scale
# that doesn't conflict with other color scales already in use.
ggnewscale_color <- function(pts, target_crs) {
  list(
    geom_sf(data = pts, aes(shape = label),
      size = 5, color = "black", fill = "white", stroke = 1.4,
      inherit.aes = FALSE),
    scale_shape_manual(
      values = c("Stream gage / outlet" = 17),
      name = NULL)
  )
}

# Mode 3: HUC12 Non-Contributing Area Choropleth
plot_choropleth <- function(da, fl, basin_label) {
  target_crs <- st_crs(3857)
  # Use the largest HUC12 set available (HUC08 superset, HUC10
  # superset, or HUC12 fallback) so every HUC12 inside the largest
  # estimated drainage area envelope appears in the choropleth.
  largest <- largest_hu12_set(da)
  hu12 <- st_transform(largest, target_crs)
  pct <- pmin(pmax(hu12$ncontrb_sqkm / hu12$dasqkm * 100, 0), 100)
  hu12$pct_noncontrib_bin <- factor(
    ifelse(pct < 33, "0-33%",
      ifelse(pct < 66, "33-66%", "66-100%")),
    levels = c("0-33%", "33-66%", "66-100%"))
  gage_pt <- get_gage_pt(da, target_crs)
  fl <- st_transform(fl, target_crs)
  basin_poly <- compute_basin_polygon(da, largest, target_crs)

  # Collect the gap-filling pieces (catchments between the lowest
  # complete HUC12 outlet and the gage) into a single sf with a
  # category column so they share one legend. The three categories
  # are distinguished by outline color and linetype; fill is uniform.
  gap_layers <- list()
  if(!is.null(da$extra_catchments) && nrow(da$extra_catchments) > 0) {
    extras <- st_transform(da$extra_catchments, target_crs)
    gap_layers$extras <- st_sf(
      gap_type = "Gap network catchment",
      geometry = st_geometry(extras))
  }
  if(!is.null(da$split_catchment) && nrow(da$split_catchment) > 0) {
    sc <- st_transform(da$split_catchment, target_crs)
    sc_full <- sc[sc$id == "catchment", ]
    sc_split <- sc[sc$id == "splitCatchment", ]
    if(nrow(sc_full) > 0)
      gap_layers$full <- st_sf(
        gap_type = "Gage local catchment",
        geometry = st_geometry(sc_full))
    if(nrow(sc_split) > 0)
      gap_layers$split <- st_sf(
        gap_type = "Split catchment at gage",
        geometry = st_geometry(sc_split))
  }
  gap_sf <- if(length(gap_layers) > 0)
    do.call(rbind, unname(gap_layers)) else NULL

  pts <- st_sf(label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt))

  p <- ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = hu12,
      aes(fill = pct_noncontrib_bin),
      color = "gray50", linewidth = 0.2, alpha = 0.8,
      inherit.aes = FALSE) +
    scale_fill_manual(
      values = c("0-33%" = "#FFFDD0",
        "33-66%" = "#C68E3F",
        "66-100%" = "#8B4513"),
      limits = c("0-33%", "33-66%", "66-100%"),
      drop = FALSE,
      name = "Non-contributing\narea")

  gap_colors <- c(
    "Gap network catchment" = "steelblue",
    "Gage local catchment" = "steelblue",
    "Split catchment at gage" = "red")
  gap_ltypes <- c(
    "Gap network catchment" = "solid",
    "Gage local catchment" = "solid",
    "Split catchment at gage" = "dashed")
  gap_lwds <- c(
    "Gap network catchment" = 0.2,
    "Gage local catchment" = 0.3,
    "Split catchment at gage" = 0.5)

  if(!is.null(gap_sf)) {
    p <- p + geom_sf(data = gap_sf,
      aes(color = gap_type, linetype = gap_type, linewidth = gap_type),
      fill = "lightblue", alpha = 0.5, inherit.aes = FALSE) +
      scale_color_manual(values = gap_colors, name = "Gap catchments") +
      scale_linetype_manual(values = gap_ltypes,
        name = "Gap catchments") +
      scale_linewidth_manual(values = gap_lwds,
        name = "Gap catchments")
  }

  p +
    geom_sf(data = fl, color = "steelblue", linewidth = 0.25,
      alpha = 0.8, inherit.aes = FALSE) +
    ggnewscale_color(pts, target_crs) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label,
      "-- HUC12 Non-Contributing Area")) +
    map_theme()
}

# Per-basin summary table
basin_summary_table <- function(da) {
  vals <- data.frame(
    source = c("Network", "HUC12", "HUC12 contributing",
      "HUC10", "HUC08", "NHDPlusHR"),
    total_sqkm = c(
      da$network_da_sqkm,
      da$da_huc12_sqkm,
      da$contrib_da_huc12_sqkm,
      ifelse(is.na(da$da_huc10_sqkm), NA_real_, da$da_huc10_sqkm),
      ifelse(is.na(da$da_huc08_sqkm), NA_real_, da$da_huc08_sqkm),
      ifelse(is.na(da$nhdplushr_network_dasqkm), NA_real_,
        da$nhdplushr_network_dasqkm)
    )
  )
  knitr::kable(vals, digits = 1,
    col.names = c("Source", "Area (sq km)"),
    caption = "Drainage area estimates by source")
}
```

## Basin Vignettes {.tabset}

### French Broad River at Asheville

Well-determined humid basin (USGS 03451500). Contributing area
essentially equals total drainage area across all sources.

#### Drainage Area Boundaries

```{r fb_boundaries}
plot_boundaries(da_results$french_broad, "French Broad")
```

```{r fb_table}
basin_summary_table(da_results$french_broad)
```

::: {.vignette-note}
- All boundary sources converge tightly around the same watershed outline.
- The HUC12-, HUC10-, and HUC08-derived boundaries are nearly identical,
  consistent with a basin whose divides are topographically unambiguous.
- The gage (triangle) sits on the main stem French Broad River at
  Asheville.
:::

#### Stream Network

```{r fb_network}
plot_network(da_results$french_broad, flowlines$french_broad,
  "French Broad")
```

::: {.vignette-note}
- The network is predominantly perennial with dense tributary coverage
  in the Appalachian headwaters.
- Intermittent reaches are sparse, concentrated in low-order headwater
  channels.
- HUC12 outlets (x markers) align with major tributary junctions.
:::

#### HUC12 Non-Contributing Area

```{r fb_choropleth, fig.height = 10}
plot_choropleth(da_results$french_broad, flowlines$french_broad,
  "French Broad")
```

::: {.vignette-note}
- Non-contributing area is negligible across all HUC12 units.
- The choropleth reads as nearly uniform light fill, confirming that
  the contributing fraction equals or nearly equals total area
  throughout the basin.
:::

### Brazos River, West Texas

Arid/ephemeral connectivity basin (USGS 08116650). Transmission
losses and disconnected uplands create large differences between total
and contributing drainage area.

#### Drainage Area Boundaries

```{r bz_boundaries}
plot_boundaries(da_results$brazos, "Brazos at Rosharon")
```

```{r bz_table}
basin_summary_table(da_results$brazos)
```

::: {.vignette-note}
- Boundary sources diverge substantially.  HUC-derived boundaries
  extend further west into arid uplands than the network-derived
  boundary.
- The gap between HUC12 total area and contributing area reflects
  large noncontributing designations in western sub-basins.
:::

#### Stream Network

```{r bz_network}
plot_network(da_results$brazos, flowlines$brazos,
  "Brazos at Rosharon")
```

::: {.vignette-note}
- Extensive intermittent and ephemeral reaches dominate the western
  (upstream) portion of the basin.
- Perennial flow is concentrated in the lower main stem and major
  tributaries east of the Caprock Escarpment.
- The transition from ephemeral headwaters to perennial mainstem
  illustrates why contributing area is smaller than total area.
:::

#### HUC12 Non-Contributing Area

```{r bz_choropleth, fig.height = 10}
plot_choropleth(da_results$brazos, flowlines$brazos,
  "Brazos at Rosharon")
```

::: {.vignette-note}
- HUC12 units in the western part of the basin carry high
  noncontributing fractions (dark brown fill).
- Noncontributing area drops sharply east of the Caprock Escarpment
  where the flow regime becomes perennial.
- The choropleth shows that the noncontributing designation is not
  distributed uniformly but concentrated in specific arid sub-basins.
:::

### James River / Cottonwood Lake

Glacial prairie basin (USGS 06468000). Prairie potholes create a
large noncontributing fraction that varies by sub-basin.

#### Drainage Area Boundaries

```{r jm_boundaries}
plot_boundaries(da_results$james, "James River")
```

```{r jm_table}
basin_summary_table(da_results$james)
```

::: {.vignette-note}
- Boundary estimates diverge in the upper basin where glacial
  topography creates ambiguous divides.
- The HUC12-derived total area exceeds the contributing area by a
  notable margin, reflecting prairie pothole storage.
:::

#### Stream Network

```{r jm_network}
plot_network(da_results$james, flowlines$james, "James River")
```

::: {.vignette-note}
- The network includes a mix of perennial and intermittent reaches.
- Intermittent channels are common in the upper basin where glacial
  drift creates closed depressions and episodic connectivity.
- The lower main stem is perennial and well-defined.
:::

#### HUC12 Non-Contributing Area

```{r jm_choropleth, fig.height = 10}
plot_choropleth(da_results$james, flowlines$james, "James River")
```

::: {.vignette-note}
- Several HUC12 units carry high noncontributing fractions (dark brown),
  concentrated in the glacial pothole region.
- The spatial pattern of noncontributing area follows the extent of
  glacial drift and closed-basin topography.
- HUC12 units in the lower basin show negligible noncontributing area.
:::

### Black Earth Creek

Small humid basin in southern Wisconsin (USGS 05406500).
Well-determined drainage area with minimal noncontributing fraction.

#### Drainage Area Boundaries

```{r be_boundaries}
plot_boundaries(da_results$black_earth, "Black Earth Creek")
```

```{r be_table}
basin_summary_table(da_results$black_earth)
```

::: {.vignette-note}
- All boundary sources converge on a compact watershed outline.
- The basin is small enough to fall within a single HUC10, so HUC10-
  and HUC08-level boundaries are not computed separately.
:::

#### Stream Network

```{r be_network}
plot_network(da_results$black_earth, flowlines$black_earth,
  "Black Earth Creek")
```

::: {.vignette-note}
- A short, predominantly perennial network drains the Driftless Area
  landscape.
- Few intermittent headwater reaches are present.
:::

#### HUC12 Non-Contributing Area

```{r be_choropleth, fig.height = 10}
plot_choropleth(da_results$black_earth, flowlines$black_earth,
  "Black Earth Creek")
```

::: {.vignette-note}
- Noncontributing area is negligible across all HUC12 units.
- The uniform light fill confirms that essentially the entire basin
  contributes to streamflow at the outlet.
:::

### Malheur Lake

Harney Basin / Malheur Lake is an endorheic (closed) basin with no
surface outlet.  The terminal feature is a lake rather than a stream
gage.

#### Drainage Area Boundaries

```{r ml_boundaries}
plot_boundaries(da_results$malheur, "Malheur Lake")
```

```{r ml_table}
basin_summary_table(da_results$malheur)
```

::: {.vignette-note}
- The basin is endorheic -- all boundaries terminate at Malheur Lake
  with no downstream outlet.
- HUC-derived and network-derived boundaries diverge in the
  surrounding high-desert uplands where surface connectivity is
  ambiguous.
- The start feature (triangle) marks the centroid of the Malheur Lake
  waterbody polygon rather than a stream gage.
:::

#### Stream Network

```{r ml_network}
plot_network(da_results$malheur, flowlines$malheur, "Malheur Lake")
```

::: {.vignette-note}
- Intermittent and ephemeral reaches dominate the network, particularly
  in the southern and eastern tributaries.
- Perennial flow is limited to the Silvies River and Donner und Blitzen
  River corridors draining into Malheur and Harney Lakes.
- The network terminates at the lake with no surface outlet downstream.
:::

#### HUC12 Non-Contributing Area

```{r ml_choropleth, fig.height = 10}
plot_choropleth(da_results$malheur, flowlines$malheur, "Malheur Lake")
```

::: {.vignette-note}
- Several HUC12 units carry high noncontributing fractions in the
  playa and alkali flat margins surrounding the lakes.
- The endorheic nature of the basin means the entire watershed is
  effectively noncontributing to any downstream system.
- The choropleth distinguishes HUC-level noncontributing designations
  from the basin-scale closed-basin classification.
:::

```{r teardown, include = FALSE}
options(oldoption)
```
