---
title: "Drainage Area Estimation"
author: "dblodgett@usgs.gov"
output:
  rmarkdown::html_vignette:
    tabset: true
vignette: >
  %\VignetteIndexEntry{Drainage Area Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo = FALSE}
/* Override html_vignette narrow body to allow wider figures */
body, div.main-container, div.body, div.contents {
  max-width: 1600px !important;
  width: 95% !important;
}
img {
  max-width: 100% !important;
  height: auto !important;
}
.vignette-note {
  background: #f8f8f8;
  border-left: 3px solid #ccc;
  padding: 8px 12px;
  margin: 8px 0 16px 0;
  font-size: 0.9em;
}
```

```{r setup, include = FALSE}
library(nhdplusTools)

build_local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 16,
  fig.height = 9,
  dpi = 110,
  out.width = "100%",
  eval = build_local
)

huc12 <- NULL
f <- "../../Data/nhdp/NHDPlusNationalData/NHDPlusV21_National_Seamless_Flattened_Lower48.gdb/"
if(file.exists(f))
  huc12 <- sf::read_sf(f, "HUC12")

has_davidson <- FALSE
oldoption <- options(scipen = 9999)
```

`get_drainage_area_estimates()` combines WBD delineations and non-surface-contributing area estimates with NHDPlusV2 catchment areas for the gap between a basin outlet
and the nearest upstream HUC12. Non-contributing areas captured in
HUC12 boundaries are included.

This vignette shows output of the function on five basins that span a range of
hydrologic settings -- from a well-determined humid basin to arid
systems, glacial prairie, and an endorheic (closed) basin.

## How `get_drainage_area_estimates()` Works

The function produces drainage area estimates by stitching together
two kinds of spatial data: HUC12 polygon areas for the bulk of the
upstream basin and NHDPlusV2 catchment areas for the gap between the
gage (or other outlet) and the nearest upstream HUC12 boundaries.
Because HUC12 polygons carry a non-contributing area attribute
(`ncontrb_a`) and non-surface-contributing HUC12s can be retrieved from HUC10 or HUC08 groupings, the estimates separate total drainage area from surface network contributing drainage area.

### Algorithm Steps

1. **Resolve the start feature.** An NLDI feature list (e.g.
   `featureSource = "nwissite", featureID = "USGS-05406500"`) is
   resolved to an NHDPlusV2 COMID via the NLDI. An `sfc_POINT` inside
   a waterbody triggers a waterbody lookup to find all outlet flowlines.

2. **Negotiate the outlet catchment.** For gage-based starts, the
   function computes where the gage sits along its outlet flowline as a
   0--100 measure. This measure determines whether the outlet catchment
   needs to be split (see *Catchment splitting* below).

3. **Fetch the upstream network and HUC12 pour points.** The full
   upstream flowline network is retrieved along with all HUC12 pour
   points on that network. These pour points mark the downstream outlet
   of each HUC12 watershed unit that drains to the gage.

4. **Identify immediate HUC12 outlets.** A network navigation
   finds which HUC12 outlets are directly upstream of the gage with no
   intervening HUC12 outlet. These define the boundary between the
   "HUC zone" (covered by HUC12 polygon areas) and the "gap zone"
   (covered by individual catchment areas).

5. **Fetch and assemble HUC12 polygons.** HUC12 polygons are retrieved
   at three query levels: the specific NLDI-identified HUC12 IDs, all
   HUC12s within upstream HUC10 boundaries, and all HUC12s within
   upstream HUC08 boundaries. The broader queries can capture
   HUC12s that share a parent HUC but lack an on-network pour point --
   common in prairie-pothole and playa-dominated landscapes. A
   disconnect filter removes HUC12s pulled incidentally by the broader
   queries that are not hydrologically connected. Each level is a
   superset of the narrower level, and each produces both a total and a
   contributing-only area estimate.

6. **Compute gap area.** Catchment areas between the gage and the
   immediate HUC12 outlets are summed. Split-catchment logic is applied
   at both the HUC12 outlets and (optionally) the gage point.

7. **Assemble scalar estimates.** Six drainage area estimates are
   computed: total and contributing at the HUC12, HUC10, and HUC08
   query levels. Each equals the HUC12 polygon area sum plus the gap
   area. A seventh scalar, `network_da_sqkm`, is the NHDPlusV2
   cumulative drainage area at the outlet -- a comparison baseline.

8. **NHDPlusHR estimate (optional).** When `nhdplushr = TRUE`, the
   function fetches NHDPlusHR flowlines and catchments for the basin
   bounding box, indexes the start point to the HR network
   (disambiguating by drainage area), navigates upstream, and sums
   catchment areas.

### Catchment Splitting

When a gage sits partway along a flowline rather than at its outlet,
the downstream portion of the catchment does not contribute to the
gage. The function calls the NLDI split-catchment service to divide
the catchment at the gage point and counts only the upstream portion.
The `outlet_split_threshold_m` parameter (default 100 m) sets the
minimum gage-to-outlet distance before splitting is performed; if the
gage is closer than this threshold, the full catchment is used.

A parallel split occurs at HUC12 outlets: the HUC boundary cuts
through a catchment, and the portion upstream of the pour point
overlaps with HUC12 area already counted. The split-catchment service
divides this catchment and the upstream overlap is excluded from the
gap area to avoid double-counting.

The *Outlet Catchment Splitting* section later in this vignette
illustrates both splits using Davidson Creek (USGS-08110075).

### Running with Local Data

Two parameters reduce dependence on web services:

- `local_navigation = TRUE` loads the NHDPlusV2 Value Added Attributes
  via `get_vaa()` for network navigation and flowline attribute
  lookups. Only HUC12 pour points are still fetched from the NLDI.

- `huc12_data` accepts a pre-loaded `sf` data.frame of HUC12 polygons
  (e.g. from the NHDPlusV2 national geodatabase). When provided, all
  HUC12 polygon queries are resolved by subsetting this table locally
  instead of calling WBD web services. The table must include `huc_12`
  and `ncontrb_a` columns (case-insensitive).

Using both together eliminates all HUC polygon web service calls and
most flowline attribute calls, leaving only the NLDI (for HUC12 pour
points and start feature resolution) and the split-catchment service.

### Web Services and Performance

By default the function contacts several web services. Each adds
latency and is subject to rate limits or outages:

- **NLDI** (`findNLDI`): resolves the start feature, navigates the
  upstream network, and retrieves HUC12 pour points. Called in every
  invocation.

- **NHDPlusV2 OGC API** (`get_nhdplus`): fetches flowline attributes,
  catchment geometries in the gap zone, and (when `catchments = TRUE`)
  the full upstream catchment polygon set. Skipped for flowline
  attributes when `local_navigation = TRUE`.

- **WBD / HUC service** (`get_huc`, `get_huc12_by_huc`): fetches
  HUC12 polygons by ID or by parent HUC. Skipped when `huc12_data` is
  provided.

- **Split-catchment service** (`get_split_catchment`): divides
  catchments at HUC12 outlets and at the gage point. Always called.

- **NHDPlusHR service** (`get_nhdphr`): fetches high-resolution
  flowlines and catchments for the basin bounding box. Skipped when
  `nhdplushr = FALSE`.

For large basins (e.g. the Brazos at Rosharon) the cumulative time for
these calls can be substantial. Setting `nhdplushr = FALSE` eliminates
the most expensive single call. Providing `huc12_data` and using
`local_navigation = TRUE` together can reduce total run time to a
fraction of the default, at the cost of requiring local data files.
The vignette's `fetch_or_load` wrapper caches results as RDS files so
that repeated runs avoid repeating the web service calls entirely.

## Fetch and Cache Results

Each basin result is stored in a separate RDS file in the nhdplusTools
data directory so that subsequent runs skip the web service calls.

```{r fetch_data, message = TRUE}
library(sf)

data_dir <- nhdplusTools_data_dir()
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

# Define the six study sites
sites <- list(
  black_earth = list(featureSource = "nwissite",
    featureID = "USGS-05406500"),
  french_broad = list(featureSource = "nwissite",
    featureID = "USGS-03451500"),
  brazos = list(featureSource = "nwissite",
    featureID = "USGS-08116650"),
  james = list(featureSource = "nwissite",
    featureID = "USGS-06468000"),
  malheur = st_sfc(st_point(c(-118.8, 43.3)), crs = 4326),
  davidson = list(featureSource = "nwissite",
    featureID = "USGS-08110075")
)

fetch_or_load <- function(name, start) {
  rds_path <- file.path(data_dir,
    paste0("da_est_", name, ".rds"))
  if(file.exists(rds_path)) {
    message("Loading cached: ", name)
    return(readRDS(rds_path))
  }
  message("Fetching: ", name)
  result <- tryCatch(
    get_drainage_area_estimates(start, catchments = TRUE,
      huc12_data = huc12),
    error = function(e) {
      warning("Failed for ", name, ": ", conditionMessage(e),
        call. = FALSE)
      NULL
    })
  if(!is.null(result)) saveRDS(result, rds_path)
  result
}

da_results <- Map(fetch_or_load, names(sites), sites)
da_results <- Filter(Negate(is.null), da_results)
```

```{r fetch_flowlines, message = TRUE}
fetch_flowlines <- function(name, da_result) {
  rds_path <- file.path(data_dir, paste0("fl_", name, ".rds"))
  if(file.exists(rds_path)) {
    message("Loading cached flowlines: ", name)
    return(readRDS(rds_path))
  }
  message("Fetching flowlines: ", name)
  comids <- da_result$all_network$comid
  fl <- tryCatch(
    get_nhdplus(comid = comids, realization = "flowline"),
    error = function(e) {
      warning("Flowline fetch failed for ", name, ": ",
        conditionMessage(e), call. = FALSE)
      NULL
    })
  if(!is.null(fl)) saveRDS(fl, rds_path)
  fl
}

flowlines <- Map(fetch_flowlines, names(da_results), da_results)
flowlines <- Filter(Negate(is.null), flowlines)
```

## Return Structure

Each result is a list with scalar drainage area estimates and spatial
data frames. Here are the elements for Black Earth Creek:

```{r structure}
names(da_results$black_earth)
```

The scalar estimates (square kilometers) across all basins:

```{r summary_table}
basin_labels <- c(
  black_earth = "Black Earth Creek",
  french_broad = "French Broad",
  brazos = "Brazos at Rosharon",
  james = "James River",
  malheur = "Malheur Lake",
  davidson = "Davidson Creek"
)

summary_df <- data.frame(
  basin = basin_labels[names(da_results)],
  network_da = vapply(da_results, \(x) x$network_da_sqkm, numeric(1)),
  da_huc12 = vapply(da_results, \(x) x$da_huc12_sqkm, numeric(1)),
  contrib_huc12 = vapply(da_results,
    \(x) x$contrib_da_huc12_sqkm, numeric(1)),
  da_huc10 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc10_sqkm), NA_real_, x$da_huc10_sqkm),
    numeric(1)),
  da_huc08 = vapply(da_results,
    \(x) ifelse(is.na(x$da_huc08_sqkm), NA_real_, x$da_huc08_sqkm),
    numeric(1)),
  nhdplushr = vapply(da_results,
    \(x) ifelse(is.na(x$nhdplushr_network_dasqkm), NA_real_,
      x$nhdplushr_network_dasqkm),
    numeric(1))
)

knitr::kable(summary_df, digits = 1,
  col.names = c("Basin", "Network DA", "HUC12 DA",
    "Contributing DA", "HUC10 DA", "HUC08 DA", "NHDPlusHR DA"),
  caption = "Drainage area estimates (sq km) by basin and method")
```

```{r helpers, include = FALSE}
library(ggplot2)

# USGS USTopo provider for maptiles. The "?f=jpg" suffix is a no-op
# query parameter that gives maptiles a recognized extension hint.
usgs_topo_provider <- list(
  src = "USGS_USTopo",
  q = paste0("https://basemap.nationalmap.gov/arcgis/rest/services/",
    "USGSTopo/MapServer/tile/{z}/{y}/{x}?f=jpg"),
  sub = NA,
  cit = "USGS The National Map: USGSTopo"
)

# Cache directory for tiles
tile_cache <- file.path(nhdplusTools_data_dir(), "da_v_tiles")
dir.create(tile_cache, recursive = TRUE, showWarnings = FALSE)

# Compute a target tile zoom level so the rendered tiles are sharp
# at the figure size we use. Web Mercator world width is 40075016 m
# and tile size is 256 px. The figure pixel width drives the
# resolution we want from the tile mosaic.
target_zoom <- function(sf_obj, fig_px = 1700) {
  bb <- st_bbox(sf_obj)
  span_m <- max(bb["xmax"] - bb["xmin"], bb["ymax"] - bb["ymin"])
  if(!is.finite(span_m) || span_m <= 0) return(10L)
  z <- log2(40075016 * fig_px / (256 * span_m))
  as.integer(min(max(round(z), 4L), 15L))
}

# Fetch USGS topo tiles for an sf bbox. Returns a list with a color
# matrix (suitable for annotation_raster) and the tile bounding box.
# Uses annotation_raster (not geom_raster) so the fill aesthetic
# stays free for choropleth use.
fetch_topo_raster <- function(sf_obj, zoom = NULL) {
  bb <- st_bbox(sf_obj)
  pad_x <- (bb["xmax"] - bb["xmin"]) * 0.05
  pad_y <- (bb["ymax"] - bb["ymin"]) * 0.05
  bb["xmin"] <- bb["xmin"] - pad_x
  bb["xmax"] <- bb["xmax"] + pad_x
  bb["ymin"] <- bb["ymin"] - pad_y
  bb["ymax"] <- bb["ymax"] + pad_y
  bb_sf <- st_as_sfc(bb, crs = st_crs(sf_obj))
  if(is.null(zoom))
    zoom <- target_zoom(st_transform(bb_sf, 3857))
  args <- list(
    x = bb_sf, provider = usgs_topo_provider, crop = TRUE,
    cachedir = tile_cache, zoom = zoom
  )
  tiles <- tryCatch(
    do.call(maptiles::get_tiles, args),
    error = function(e) {
      warning("Tile fetch failed: ", conditionMessage(e),
        call. = FALSE)
      NULL
    })
  if(is.null(tiles)) return(NULL)
  rgb_arr <- terra::as.array(tiles)
  if(length(dim(rgb_arr)) != 3 || dim(rgb_arr)[3] < 3) return(NULL)
  col_mat <- matrix(
    rgb(rgb_arr[, , 1], rgb_arr[, , 2], rgb_arr[, , 3],
      maxColorValue = 255),
    nrow = dim(rgb_arr)[1], ncol = dim(rgb_arr)[2]
  )
  ext <- terra::ext(tiles)
  list(
    raster = col_mat,
    xmin = ext[1], xmax = ext[2],
    ymin = ext[3], ymax = ext[4]
  )
}

# Add basemap as annotation_raster (does not consume fill scale)
add_topo <- function(p, sf_obj) {
  topo <- fetch_topo_raster(sf_obj)
  if(is.null(topo)) return(p)
  p + annotation_raster(
    topo$raster,
    xmin = topo$xmin, xmax = topo$xmax,
    ymin = topo$ymin, ymax = topo$ymax
  )
}

# Resolve gage point in target CRS
get_gage_pt <- function(da, target_crs) {
  pt <- st_geometry(da$start_feature)
  if(!inherits(pt, "sfc_POINT"))
    pt <- st_centroid(pt)
  st_transform(st_sf(geometry = pt), target_crs)
}

# Pick the spatially-largest of the available HUC12 sets returned by
# get_drainage_area_estimates: hu12_by_huc08 (if present), otherwise
# hu12_by_huc10, otherwise hu12_by_huc12. Largest is determined by
# the sum of HUC12 dasqkm in each set.
largest_hu12_set <- function(da) {
  candidates <- list(
    huc12 = da$hu12_by_huc12,
    huc10 = da$hu12_by_huc10,
    huc08 = da$hu12_by_huc08
  )
  candidates <- Filter(
    function(x) !is.null(x) && nrow(x) > 0, candidates)
  areas <- vapply(candidates,
    function(x) sum(x$dasqkm, na.rm = TRUE), numeric(1))
  candidates[[which.max(areas)]]
}

# Compute a single merged basin polygon for a given HUC level set
# (hu12_by_huc12, hu12_by_huc10, or hu12_by_huc08), unioned with
# extra_catchments and the local catchments around the gage.
# Uses hydroloom::dissolve_polygons() with gap_tolerance to absorb
# sliver gaps between polygons from different services (WBD HUC12s,
# NHDPlusV2 catchments, split-catchment service).
compute_basin_polygon <- function(da, hu_layer, target_crs,
  tol_m = 250) {
  common_crs <- st_crs(hu_layer)
  pieces <- list()
  pieces$hu <- st_geometry(hu_layer)

  if(!is.null(da$extra_catchments) && nrow(da$extra_catchments) > 0)
    pieces$extra <- st_geometry(
      st_transform(da$extra_catchments, common_crs))

  if(!is.null(da$split_catchment) && nrow(da$split_catchment) > 0) {
    sc <- da$split_catchment
    sc_full <- sc[sc$id == "catchment", ]
    if(nrow(sc_full) > 0)
      pieces$catch <- st_geometry(st_transform(sc_full, common_crs))
  }

  all_polys <- st_sf(
    geometry = do.call(c, pieces)
  ) |> st_set_crs(common_crs)

  hydroloom::dissolve_polygons(all_polys,
    gap_tolerance = tol_m,
    max_hole_area = Inf,
    single_polygon = TRUE) |>
    st_transform(target_crs)
}

# Common theme: no axis ticks, labels, or grid
map_theme <- function() {
  theme_void() +
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      plot.title = element_text(hjust = 0, size = 13,
        margin = margin(b = 6))
    )
}

# Add the gray basin underlay to a plot. The polygon represents the
# full estimated drainage area envelope (largest available HUC level
# plus extra catchments and local catchments).
add_basin_underlay <- function(p, basin_poly) {
  p + geom_sf(data = basin_poly,
    fill = "gray30", color = NA, alpha = 0.15,
    inherit.aes = FALSE)
}

# Mode 1: Drainage Area Boundaries
plot_boundaries <- function(da, basin_label) {
  # Reproject working layers to Web Mercator (3857) so the basemap
  # tiles align cleanly.
  target_crs <- st_crs(3857)
  gage_pt <- get_gage_pt(da, target_crs)
  basin_poly <- compute_basin_polygon(da, largest_hu12_set(da), target_crs)

  # Dissolve boundaries at each HUC level. Each HUC level is combined
  # with extra_catchments and split_catchment via compute_basin_polygon
  # so every boundary reaches the gage instead of stopping at the
  # lowest complete HUC outlet.
  boundaries <- list()
  boundaries[["HUC12 (NLDI)"]] <- st_geometry(
    compute_basin_polygon(da, da$hu12_by_huc12, target_crs))
  if(!is.null(da$hu12_by_huc10))
    boundaries[["HUC10"]] <- st_geometry(
      compute_basin_polygon(da, da$hu12_by_huc10, target_crs))
  if(!is.null(da$hu12_by_huc08))
    boundaries[["HUC08"]] <- st_geometry(
      compute_basin_polygon(da, da$hu12_by_huc08, target_crs))
  if(!is.null(da$all_catchments))
    boundaries[["Network catchments"]] <- st_union(
      st_transform(da$all_catchments, target_crs))

  ltypes <- c("solid", "dashed", "dotted", "twodash")

  bnd_sf <- do.call(rbind, lapply(seq_along(boundaries), function(i) {
    st_sf(source = names(boundaries)[i],
      geometry = boundaries[[i]])
  }))

  hu12_pp <- if(!is.null(da$hu12_outlet))
    st_transform(da$hu12_outlet, target_crs) else NULL

  # Combine point layers into one sf so a shape scale can label both
  pts <- st_sf(
    label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt)
  )
  if(!is.null(hu12_pp)) {
    pts <- rbind(
      pts,
      st_sf(label = "HUC12 outlet",
        geometry = st_geometry(hu12_pp))
    )
  }

  p <- ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = bnd_sf, aes(linetype = source),
      fill = NA, color = "gray10", linewidth = 0.8,
      inherit.aes = FALSE) +
    scale_linetype_manual(
      values = setNames(ltypes[seq_along(boundaries)],
        names(boundaries)),
      name = "DA boundary") +
    geom_sf(data = pts, aes(shape = label, size = label,
      color = label),
      fill = "white", stroke = 1.4, inherit.aes = FALSE) +
    scale_shape_manual(
      values = c("Stream gage / outlet" = 17, "HUC12 outlet" = 4),
      name = NULL) +
    scale_size_manual(
      values = c("Stream gage / outlet" = 5, "HUC12 outlet" = 3),
      name = NULL) +
    scale_color_manual(
      values = c("Stream gage / outlet" = "black",
        "HUC12 outlet" = "darkred"),
      name = NULL) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label, "-- Drainage Area Boundaries")) +
    map_theme()

  p
}

# Mode 2: Stream Network
plot_network <- function(da, fl, basin_label) {
  target_crs <- st_crs(3857)
  gage_pt <- get_gage_pt(da, target_crs)
  fl <- st_transform(fl, target_crs)
  basin_poly <- compute_basin_polygon(da, largest_hu12_set(da), target_crs)

  # fcode 46003 = intermittent, 46007 = ephemeral, 46006 = perennial
  fl$flow_class <- ifelse(
    fl$fcode %in% c(46003L, 46007L), "Intermittent/Ephemeral",
    "Perennial/Other"
  )
  fl$lwd <- pmin(pmax(fl$streamorde, 1) / 3, 1.5)

  pts <- st_sf(label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt))

  ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = fl,
      aes(color = flow_class, linewidth = lwd),
      show.legend = "line", inherit.aes = FALSE) +
    scale_color_manual(
      values = c("Perennial/Other" = "steelblue",
        "Intermittent/Ephemeral" = "lightskyblue"),
      name = "Flow class") +
    scale_linewidth_identity() +
    ggnewscale_color(pts, target_crs) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label, "-- Stream Network")) +
    map_theme()
}

# Helper to add the gage point as a labeled layer using a shape scale
# that doesn't conflict with other color scales already in use.
ggnewscale_color <- function(pts, target_crs) {
  list(
    geom_sf(data = pts, aes(shape = label),
      size = 5, color = "black", fill = "white", stroke = 1.4,
      inherit.aes = FALSE),
    scale_shape_manual(
      values = c("Stream gage / outlet" = 17),
      name = NULL)
  )
}

# Mode 3: HUC12 Non-Contributing Area Choropleth
plot_choropleth <- function(da, fl, basin_label) {
  target_crs <- st_crs(3857)
  # Use the largest HUC12 set available (HUC08 superset, HUC10
  # superset, or HUC12 fallback) so every HUC12 inside the largest
  # estimated drainage area envelope appears in the choropleth.
  largest <- largest_hu12_set(da)
  hu12 <- st_transform(largest, target_crs)
  pct <- pmin(pmax(hu12$ncontrb_sqkm / hu12$dasqkm * 100, 0), 100)
  hu12$pct_noncontrib_bin <- factor(
    ifelse(pct < 33, "0-33%",
      ifelse(pct < 66, "33-66%", "66-100%")),
    levels = c("0-33%", "33-66%", "66-100%"))
  gage_pt <- get_gage_pt(da, target_crs)
  fl <- st_transform(fl, target_crs)
  basin_poly <- compute_basin_polygon(da, largest, target_crs)

  # Collect the gap-filling pieces (catchments between the lowest
  # complete HUC12 outlet and the gage) into a single sf with a
  # category column so they share one legend. The three categories
  # are distinguished by outline color and linetype; fill is uniform.
  gap_layers <- list()
  if(!is.null(da$extra_catchments) && nrow(da$extra_catchments) > 0) {
    extras <- st_transform(da$extra_catchments, target_crs)
    gap_layers$extras <- st_sf(
      gap_type = "Gap network catchment",
      geometry = st_geometry(extras))
  }
  if(!is.null(da$split_catchment) && nrow(da$split_catchment) > 0) {
    sc <- st_transform(da$split_catchment, target_crs)
    sc_full <- sc[sc$id == "catchment", ]
    sc_split <- sc[sc$id == "splitCatchment", ]
    if(nrow(sc_full) > 0)
      gap_layers$full <- st_sf(
        gap_type = "Gage local catchment",
        geometry = st_geometry(sc_full))
    if(nrow(sc_split) > 0)
      gap_layers$split <- st_sf(
        gap_type = "Split catchment at gage",
        geometry = st_geometry(sc_split))
  }
  gap_sf <- if(length(gap_layers) > 0)
    do.call(rbind, unname(gap_layers)) else NULL

  pts <- st_sf(label = "Stream gage / outlet",
    geometry = st_geometry(gage_pt))

  p <- ggplot() |>
    add_topo(basin_poly) |>
    add_basin_underlay(basin_poly) +
    geom_sf(data = hu12,
      aes(fill = pct_noncontrib_bin),
      color = "gray50", linewidth = 0.2, alpha = 0.8,
      inherit.aes = FALSE) +
    scale_fill_manual(
      values = c("0-33%" = "#FFFDD0",
        "33-66%" = "#C68E3F",
        "66-100%" = "#8B4513"),
      limits = c("0-33%", "33-66%", "66-100%"),
      drop = FALSE,
      name = "Non-contributing\narea")

  gap_colors <- c(
    "Gap network catchment" = "steelblue",
    "Gage local catchment" = "steelblue",
    "Split catchment at gage" = "red")
  gap_ltypes <- c(
    "Gap network catchment" = "solid",
    "Gage local catchment" = "solid",
    "Split catchment at gage" = "dashed")
  gap_lwds <- c(
    "Gap network catchment" = 0.2,
    "Gage local catchment" = 0.3,
    "Split catchment at gage" = 0.5)

  if(!is.null(gap_sf)) {
    p <- p + geom_sf(data = gap_sf,
      aes(color = gap_type, linetype = gap_type, linewidth = gap_type),
      fill = "lightblue", alpha = 0.5, inherit.aes = FALSE) +
      scale_color_manual(values = gap_colors, name = "Gap catchments") +
      scale_linetype_manual(values = gap_ltypes,
        name = "Gap catchments") +
      scale_linewidth_manual(values = gap_lwds,
        name = "Gap catchments")
  }

  p +
    geom_sf(data = fl, color = "steelblue", linewidth = 0.25,
      alpha = 0.8, inherit.aes = FALSE) +
    ggnewscale_color(pts, target_crs) +
    coord_sf(crs = target_crs, expand = FALSE) +
    labs(title = paste(basin_label,
      "-- HUC12 Non-Contributing Area")) +
    map_theme()
}

# Per-basin summary table
basin_summary_table <- function(da) {
  vals <- data.frame(
    source = c("Network", "HUC12", "HUC12 contributing",
      "HUC10", "HUC08", "NHDPlusHR"),
    total_sqkm = c(
      da$network_da_sqkm,
      da$da_huc12_sqkm,
      da$contrib_da_huc12_sqkm,
      ifelse(is.na(da$da_huc10_sqkm), NA_real_, da$da_huc10_sqkm),
      ifelse(is.na(da$da_huc08_sqkm), NA_real_, da$da_huc08_sqkm),
      ifelse(is.na(da$nhdplushr_network_dasqkm), NA_real_,
        da$nhdplushr_network_dasqkm)
    )
  )
  knitr::kable(vals, digits = 1,
    col.names = c("Source", "Area (sq km)"),
    caption = "Drainage area estimates by source")
}
```

## Basin Vignettes {.tabset}

### French Broad River at Asheville

Well-determined humid basin (USGS 03451500). Contributing area
essentially equals total drainage area across all sources.

#### Drainage Area Boundaries

```{r fb_boundaries}
plot_boundaries(da_results$french_broad, "French Broad")
```

```{r fb_table}
basin_summary_table(da_results$french_broad)
```

::: {.vignette-note}
- All boundary sources converge tightly around the same watershed outline.
- The HUC12-, HUC10-, and HUC08-derived boundaries are nearly identical,
  consistent with a basin whose divides are topographically unambiguous.
- The gage (triangle) sits on the main stem French Broad River at
  Asheville.
:::

#### Stream Network

```{r fb_network}
plot_network(da_results$french_broad, flowlines$french_broad,
  "French Broad")
```

::: {.vignette-note}
- The network is predominantly perennial with dense tributary coverage
  in the Appalachian headwaters.
- Intermittent reaches are sparse, concentrated in low-order headwater
  channels.
- HUC12 outlets (x markers) align with major tributary junctions.
:::

#### HUC12 Non-Contributing Area

```{r fb_choropleth, fig.height = 10}
plot_choropleth(da_results$french_broad, flowlines$french_broad,
  "French Broad")
```

::: {.vignette-note}
- Non-contributing area is negligible across all HUC12 units.
- The choropleth reads as nearly uniform light fill, confirming that
  the contributing fraction equals or nearly equals total area
  throughout the basin.
:::

### Brazos River, West Texas

Arid/ephemeral connectivity basin (USGS 08116650). Transmission
losses and disconnected uplands create large differences between total
and contributing drainage area.

#### Drainage Area Boundaries

```{r bz_boundaries}
plot_boundaries(da_results$brazos, "Brazos at Rosharon")
```

```{r bz_table}
basin_summary_table(da_results$brazos)
```

::: {.vignette-note}
- Boundary sources diverge substantially.  HUC-derived boundaries
  extend further west into arid uplands than the network-derived
  boundary.
- The gap between HUC12 total area and contributing area reflects
  large noncontributing designations in western sub-basins.
:::

#### Stream Network

```{r bz_network}
plot_network(da_results$brazos, flowlines$brazos,
  "Brazos at Rosharon")
```

::: {.vignette-note}
- Extensive intermittent and ephemeral reaches dominate the western
  (upstream) portion of the basin.
- Perennial flow is concentrated in the lower main stem and major
  tributaries east of the Caprock Escarpment.
- The transition from ephemeral headwaters to perennial mainstem
  illustrates why contributing area is smaller than total area.
:::

#### HUC12 Non-Contributing Area

```{r bz_choropleth, fig.height = 10}
plot_choropleth(da_results$brazos, flowlines$brazos,
  "Brazos at Rosharon")
```

::: {.vignette-note}
- HUC12 units in the western part of the basin carry high
  noncontributing fractions (dark brown fill).
- Noncontributing area drops sharply east of the Caprock Escarpment
  where the flow regime becomes perennial.
- The choropleth shows that the noncontributing designation is not
  distributed uniformly but concentrated in specific arid sub-basins.
:::

### James River / Cottonwood Lake

Glacial prairie basin (USGS 06468000). Prairie potholes create a
large noncontributing fraction that varies by sub-basin.

#### Drainage Area Boundaries

```{r jm_boundaries}
plot_boundaries(da_results$james, "James River")
```

```{r jm_table}
basin_summary_table(da_results$james)
```

::: {.vignette-note}
- Boundary estimates diverge in the upper basin where glacial
  topography creates ambiguous divides.
- The HUC12-derived total area exceeds the contributing area by a
  notable margin, reflecting prairie pothole storage.
:::

#### Stream Network

```{r jm_network}
plot_network(da_results$james, flowlines$james, "James River")
```

::: {.vignette-note}
- The network includes a mix of perennial and intermittent reaches.
- Intermittent channels are common in the upper basin where glacial
  drift creates closed depressions and episodic connectivity.
- The lower main stem is perennial and well-defined.
:::

#### HUC12 Non-Contributing Area

```{r jm_choropleth, fig.height = 10}
plot_choropleth(da_results$james, flowlines$james, "James River")
```

::: {.vignette-note}
- Several HUC12 units carry high noncontributing fractions (dark brown),
  concentrated in the glacial pothole region.
- The spatial pattern of noncontributing area follows the extent of
  glacial drift and closed-basin topography.
- HUC12 units in the lower basin show negligible noncontributing area.
:::

### Black Earth Creek

Small humid basin in southern Wisconsin (USGS 05406500).
Well-determined drainage area with minimal noncontributing fraction.

#### Drainage Area Boundaries

```{r be_boundaries}
plot_boundaries(da_results$black_earth, "Black Earth Creek")
```

```{r be_table}
basin_summary_table(da_results$black_earth)
```

::: {.vignette-note}
- All boundary sources converge on a compact watershed outline.
- The basin is small enough to fall within a single HUC10, so HUC10-
  and HUC08-level boundaries are not computed separately.
:::

#### Stream Network

```{r be_network}
plot_network(da_results$black_earth, flowlines$black_earth,
  "Black Earth Creek")
```

::: {.vignette-note}
- A short, predominantly perennial network drains the Driftless Area
  landscape.
- Few intermittent headwater reaches are present.
:::

#### HUC12 Non-Contributing Area

```{r be_choropleth, fig.height = 10}
plot_choropleth(da_results$black_earth, flowlines$black_earth,
  "Black Earth Creek")
```

::: {.vignette-note}
- Noncontributing area is negligible across all HUC12 units.
- The uniform light fill confirms that essentially the entire basin
  contributes to streamflow at the outlet.
:::

### Malheur Lake

Harney Basin / Malheur Lake is an endorheic (closed) basin with no
surface outlet.  The terminal feature is a lake rather than a stream
gage.

#### Drainage Area Boundaries

```{r ml_boundaries}
plot_boundaries(da_results$malheur, "Malheur Lake")
```

```{r ml_table}
basin_summary_table(da_results$malheur)
```

::: {.vignette-note}
- The basin is endorheic -- all boundaries terminate at Malheur Lake
  with no downstream outlet.
- HUC-derived and network-derived boundaries diverge in the
  surrounding high-desert uplands where surface connectivity is
  ambiguous.
- The start feature (triangle) marks the centroid of the Malheur Lake
  waterbody polygon rather than a stream gage.
:::

#### Stream Network

```{r ml_network}
plot_network(da_results$malheur, flowlines$malheur, "Malheur Lake")
```

::: {.vignette-note}
- Intermittent and ephemeral reaches dominate the network, particularly
  in the southern and eastern tributaries.
- Perennial flow is limited to the Silvies River and Donner und Blitzen
  River corridors draining into Malheur and Harney Lakes.
- The network terminates at the lake with no surface outlet downstream.
:::

#### HUC12 Non-Contributing Area

```{r ml_choropleth, fig.height = 10}
plot_choropleth(da_results$malheur, flowlines$malheur, "Malheur Lake")
```

::: {.vignette-note}
- Several HUC12 units carry high noncontributing fractions in the
  playa and alkali flat margins surrounding the lakes.
- The endorheic nature of the basin means the entire watershed is
  effectively noncontributing to any downstream system.
- The choropleth distinguishes HUC-level noncontributing designations
  from the basin-scale closed-basin classification.
:::

## Outlet Catchment Splitting

When a gage sits partway along a flowline rather than at its outlet,
the gage's outlet catchment is only partly upstream of the gage. The
downstream portion does not contribute to the gage and should be
excluded. A similar situation arises at HUC12 outlets: the HUC
boundary cuts through a catchment and the portion upstream of the
pour point overlaps with the HUC area that is already counted.

`get_drainage_area_estimates()` handles both splits automatically.
The `outlet_split_threshold_m` parameter (default 100 m) controls the
minimum gage-to-outlet distance before the split is performed.

This example uses a small Texas gage (USGS-08110075, Davidson Creek)
where the gage falls roughly two thirds of the way up its flowline.

```{r outlet_split_prep, include = FALSE}
da_dav <- da_results$davidson
fl_dav <- flowlines$davidson
has_davidson <- !is.null(da_dav) && !is.null(fl_dav) &&
  !is.null(da_dav$outlet_split_catchment)

if(has_davidson) {
  target_crs <- st_crs(3857)

  gage_pt <- get_gage_pt(da_dav, target_crs)
  gage_comid <- as.integer(da_dav$start_feature$comid)

  # all catchments in the gap between the gage and HUC12 outlets
  extra_cat <- st_transform(da_dav$extra_catchments, target_crs)
  all_cat <- st_transform(da_dav$all_catchments, target_crs)
  fl_sf <- st_transform(fl_dav, target_crs)
  hu12_pts <- st_transform(da_dav$hu12_outlet, target_crs)

  # HUC12 split catchment (split at the HUC12 pour point)
  hu12_sc <- st_transform(da_dav$split_catchment, target_crs)
  hu12_catch_full <- hu12_sc[hu12_sc$id == "catchment", ]
  hu12_catch_split <- hu12_sc[hu12_sc$id == "splitCatchment", ]
  hu12_comid <- as.integer(na.omit(hu12_sc$catchmentID))

  # gage outlet split catchment (split at the gage point)
  osc <- st_transform(da_dav$outlet_split_catchment, target_crs)
  osc_full <- osc[osc$id == "catchment", ]
  osc_split <- osc[osc$id == "splitCatchment", ]

  # flowlines in the gap area
  gap_comids <- c(gage_comid, extra_cat$featureid, hu12_comid)
  gap_fl <- fl_sf[fl_sf$comid %in% gap_comids, ]

  # focus bbox: union of gap catchments + split catchments + gage
  focus_geom <- st_union(c(
    st_geometry(extra_cat),
    st_geometry(hu12_catch_full),
    st_geometry(osc_full),
    st_geometry(gage_pt)
  ))
  focus_bb <- st_bbox(focus_geom)
  pad_x <- (focus_bb["xmax"] - focus_bb["xmin"]) * 0.05
  pad_y <- (focus_bb["ymax"] - focus_bb["ymin"]) * 0.05
  focus_xlim <- c(focus_bb["xmin"] - pad_x, focus_bb["xmax"] + pad_x)
  focus_ylim <- c(focus_bb["ymin"] - pad_y, focus_bb["ymax"] + pad_y)
}
```

#### Overview: Gage and HUC Outlet Positions

The gage (triangle) and the nearest HUC12 outlet (x) sit on
different catchments with a gap between them. The gap catchments
(light blue) and split catchment boundaries define the area between
the gage and the HUC12 boundary.

```{r outlet_split_overview, fig.height = 10, eval = has_davidson}
p_overview <- ggplot() |>
  add_topo(focus_geom) +
  # all catchments as light underlay
  geom_sf(data = all_cat, fill = "gray30", color = NA, alpha = 0.12,
    inherit.aes = FALSE) +
  # gap catchments
  geom_sf(data = extra_cat, fill = "lightblue", color = "steelblue",
    linewidth = 0.3, alpha = 0.5, inherit.aes = FALSE) +
  # HUC12 split catchment — full outline
  geom_sf(data = hu12_catch_full, fill = NA, color = "gray10",
    linewidth = 0.6, linetype = "dashed", inherit.aes = FALSE) +
  # gage outlet catchment — full outline
  geom_sf(data = osc_full, fill = NA, color = "gray10",
    linewidth = 0.6, inherit.aes = FALSE) +
  # flowlines in the gap
  geom_sf(data = gap_fl, color = "steelblue", linewidth = 0.5,
    inherit.aes = FALSE) +
  # HUC12 outlet
  geom_sf(data = hu12_pts[hu12_pts$comid == hu12_comid, ],
    shape = 4, color = "darkred", size = 4, stroke = 1.6,
    inherit.aes = FALSE) +
  # gage point
  geom_sf(data = gage_pt, shape = 17, color = "black",
    fill = "white", size = 5, stroke = 1.4,
    inherit.aes = FALSE) +
  coord_sf(crs = target_crs, xlim = focus_xlim, ylim = focus_ylim,
    expand = FALSE) +
  labs(title = paste0("Davidson Creek (USGS-08110075)",
    " -- Gage and HUC12 Outlet Positions")) +
  map_theme()
print(p_overview)
```

::: {.vignette-note}
- The gage (triangle) and the HUC12 pour point (x) sit on different
  catchments with several gap catchments (light blue) between them.
- The dashed outline marks the catchment where the HUC12 pour point
  falls; the solid outline marks the gage's catchment.
- Both the gage and the HUC12 outlet sit partway along their
  respective flowlines, so splitting is needed at both locations.
:::

#### HUC Outlet Detail

The HUC12 split catchment is small enough that it is not visible in the
overview. This view zooms to a 500 m square centered on the HUC12 outlet.

```{r outlet_split_huc_zoom, fig.height = 10, eval = has_davidson}
hu12_outlet_pt <- hu12_pts[hu12_pts$comid == hu12_comid, ]
hu12_coords <- st_coordinates(hu12_outlet_pt)
half_side <- 250 # meters in EPSG:3857
zoom_xlim <- c(hu12_coords[1, "X"] - half_side,
  hu12_coords[1, "X"] + half_side)
zoom_ylim <- c(hu12_coords[1, "Y"] - half_side,
  hu12_coords[1, "Y"] + half_side)

p_huc_zoom <- ggplot() |>
  add_topo(hu12_outlet_pt) +
  # gap catchments
  geom_sf(data = extra_cat, fill = "lightblue", color = "steelblue",
    linewidth = 0.3, alpha = 0.5, inherit.aes = FALSE) +
  # HUC12 split catchment — full outline
  geom_sf(data = hu12_catch_full, fill = NA, color = "gray10",
    linewidth = 0.6, linetype = "dashed", inherit.aes = FALSE) +
  # HUC12 split portion
  geom_sf(data = hu12_catch_split, fill = "orange", color = "gray10",
    linewidth = 0.4, alpha = 0.5, inherit.aes = FALSE) +
  # flowlines in the gap
  geom_sf(data = gap_fl, color = "steelblue", linewidth = 0.5,
    inherit.aes = FALSE) +
  # HUC12 outlet
  geom_sf(data = hu12_outlet_pt,
    shape = 4, color = "darkred", size = 4, stroke = 1.6,
    inherit.aes = FALSE) +
  coord_sf(crs = target_crs, xlim = zoom_xlim, ylim = zoom_ylim,
    expand = FALSE) +
  labs(title = paste0("Davidson Creek -- HUC12 Outlet Detail",
    " (500 m view)")) +
  map_theme()
print(p_huc_zoom)
```

::: {.vignette-note}
- The HUC12 outlet (x) and its split catchment (orange fill, dashed
  outline) are visible at this scale.
- The split catchment upstream of the HUC outlet is excluded from the
  gap area because it overlaps with the HUC12 drainage area.
:::

#### Split Catchments: Upstream and Downstream Portions

Two catchments are split. At the gage, the downstream portion is
removed because it does not contribute flow to the gage. At the
HUC12 outlet, the upstream portion is removed because it overlaps
with the HUC12 area already counted in the drainage area estimate.

```{r outlet_split_detail, fig.height = 10, eval = has_davidson}
# Build an sf with labeled polygons for a single legend
split_layers <- rbind(
  st_sf(
    role = "Upstream of gage (included)",
    geometry = st_geometry(osc_split)),
  st_sf(
    role = "Downstream of gage (excluded)",
    geometry = st_difference(
      st_geometry(osc_full), st_geometry(osc_split))),
  st_sf(
    role = "Upstream of HUC outlet (excluded, overlaps HUC)",
    geometry = st_geometry(hu12_catch_split)),
  st_sf(
    role = "Downstream of HUC outlet (included as local area)",
    geometry = st_difference(
      st_geometry(hu12_catch_full), st_geometry(hu12_catch_split)))
)

split_colors <- c(
  "Upstream of gage (included)" = "#4DAF4A",
  "Downstream of gage (excluded)" = "#E41A1C",
  "Upstream of HUC outlet (excluded, overlaps HUC)" = "#FF7F00",
  "Downstream of HUC outlet (included as local area)" = "#377EB8"
)

split_layers$role <- factor(split_layers$role,
  levels = names(split_colors))

p_split <- ggplot() |>
  add_topo(focus_geom) +
  # gap catchments as underlay
  geom_sf(data = extra_cat, fill = "gray80", color = "gray60",
    linewidth = 0.2, alpha = 0.3, inherit.aes = FALSE) +
  # split polygons with role-based fill
  geom_sf(data = split_layers,
    aes(fill = role), color = "gray20", linewidth = 0.4,
    alpha = 0.6, inherit.aes = FALSE) +
  scale_fill_manual(values = split_colors, name = NULL) +
  # flowlines
  geom_sf(data = gap_fl, color = "steelblue", linewidth = 0.5,
    inherit.aes = FALSE) +
  # HUC12 outlet
  geom_sf(data = hu12_pts[hu12_pts$comid == hu12_comid, ],
    shape = 4, color = "darkred", size = 4, stroke = 1.6,
    inherit.aes = FALSE) +
  # gage
  geom_sf(data = gage_pt, shape = 17, color = "black",
    fill = "white", size = 5, stroke = 1.4,
    inherit.aes = FALSE) +
  coord_sf(crs = target_crs, xlim = focus_xlim, ylim = focus_ylim,
    expand = FALSE) +
  labs(title = paste0("Davidson Creek -- Split Catchment Roles"),
    subtitle = paste0(
      "Flowline measure at gage: ",
      round(da_dav$outlet_flowline_measure, 1),
      "; downstream removed: ",
      round(osc_full$dasqkm - osc_split$dasqkm, 2), " km\u00B2")) +
  map_theme() +
  guides(fill = guide_legend(ncol = 1))
print(p_split)
```

::: {.vignette-note}
- **Green**: the portion of the gage's catchment upstream of the gage —
  this area is included in the drainage area estimate.
- **Red**: the portion downstream of the gage — excluded because it
  does not contribute flow to the gage.
- **Orange**: the portion of the HUC12 outlet catchment upstream of the
  pour point — excluded because this area is already counted as part
  of the HUC12 drainage area.
- **Blue**: the portion downstream of the HUC12 outlet — included as
  local area in the gap between the HUC boundary and the gage.
- The `outlet_split_threshold_m` parameter (default 100 m) controls
  whether the gage split is performed. If the gage is within the
  threshold distance of the catchment outlet, no split occurs.
:::

```{r teardown, include = FALSE}
options(oldoption)
```
