---
title: "Introduction to nhdplusTools"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nhdplusTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(nhdplusTools)

local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")
if(local) {
  cache_path <- file.path(nhdplusTools_data_dir(), "nhdpt_v_cache")
} else {
  cache_path <- tempdir()
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4,
  eval=local,
  cache=local,
  cache.path=(cache_path)
)

oldoption <- options(scipen = 9999,
                     "rgdal_show_exportToProj4_warnings"="none")

```

## TL;DR

The nhdplusTools package supports a wide range of hydrographic data use cases. Many are globally applicable hydrographic data manipulation and processing utilities. Others are specific to a U.S. context, focused on particular US datasets and sources. 

Detailed documentation of all the package functions can be found at the [Reference page](https://usgs-r.github.io/nhdplusTools/reference/) and [Articles index](https://usgs-r.github.io/nhdplusTools/articles/).

For data discovery an access in a U.S. context, start with: `vignette("US_data")`.

Globally applicable hydrographic data functions, start with the `vignette("advanced_network")`.

Please consider registering [issues and feature suggestions on github](https://github.com/dblodgett-usgs/nhdplusTools/issues).

## Installation

For the latest CRAN release of nhdplusTools, use:
```{r cran, echo = TRUE, eval = FALSE}
# install.packages("nhdplusTools")
```

The easiest way to install the latest `nhdplusTools` development is with the `remotes`` package like this:
```{r install, echo = TRUE, eval = FALSE}
# install.packages("remotes")
# remotes::install_github("usgs-r/nhdplusTools")
```

## Package Summary

The `nhdplusTools` package is intended to provide a reusable set of tools to
subset, relate data to, and rebuild attributes for hydrography data. 
It implements a data model consistent with both the [NHDPlus](https://www.epa.gov/waterdata/nhdplus-national-hydrography-dataset-plus)
and [HY\_Features](http://opengeospatial.github.io/HY_Features/). The package
aims to provide a set of tools with minimal dependencies that can be used
to build workflows using NHDPlus data-model data.

The package has three types of functionality:

1. `vignette("US_data")`
1. `vignette("point_indexing")`
1. `vignette("advanced_network")`

Two other vignette articles are available for the package.

1. `vignette("nhdplushr")`
1. `vignette("plot_nhdplus")`

Thanks for taking the time to read the package documentation. If you see anything that looks wrong or could be better, please consider opening an issue of a pull request in [github](https://github.com/dblodgett-usgs/nhdplusTools/issues)!

```{r teardown, include=FALSE}
options(oldoption)

if(Sys.getenv("BUILD_VIGNETTES") != "TRUE") {
  unlink(work_dir, recursive = TRUE)
}
```
