---
title: "Plotting with nhdplusTools"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot_nhdplus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4,
  eval=nzchar(Sys.getenv("BUILD_VIGNETTES"))
)
oldoption <- options(scipen = 9999)
options(scipen = 9999)
```
```{r data_dir_setup, echo=FALSE, include=FALSE}
temp_dir <- tempdir()
dir.create(temp_dir)
```

# Plotting with nhdplusTools

The goal of this vignette is to demonstrate a simple and lightweight approach to building maps with NHDPlus data. 

## Getting Data

For this example, we'll start from an outlet NWIS Site. Note that other options are possible with `discover_nhdplus_id` and `discover_nldi_sources`.

```{r get data}
library(sf)
library(nhdplusTools)
nwissite <- list(featureSource = "nwissite", 
                     featureID = "USGS-05428500")

flowline <- navigate_nldi(nwissite, 
                          mode = "upstreamTributaries", 
                          data_source = "")

nhdplus <- subset_nhdplus(comids = flowline$nhdplus_comid,
                          output_file = file.path(temp_dir, "nhdplus.gpkg"),
                          nhdplus_data = "download",
                          overwrite = TRUE)

flowline <- read_sf(nhdplus, "NHDFlowline_Network")

upstream_nwis <- navigate_nldi(nwissite,
                               mode = "upstreamTributaries",
                               data_source = "nwissite")

basin <- get_nldi_basin(nwissite)
```

Now we have a file at the path held in the variable `nhdplus` and three `sf` `data.frame`s with contents that look like:

```{r introspect} 
st_layers(nhdplus)
names(flowline)
names(upstream_nwis)
names(basin)
class(st_geometry(flowline))
class(st_geometry(upstream_nwis))
class(st_geometry(basin))
```

Our file has four layers: network flowlines, simplified catchments, nhd area features, and nhd waterbody features. 

The flowlines have a large set of attributes from the NHDPlus dataset. And the nwis sites have a few attributes that came from the NLDI. Attributes for NWIS sites can be found using the [dataRetrieval package.](https://owi.usgs.gov/R/dataRetrieval.html)

See the NHDPlus user guide [linked here](https://www.epa.gov/waterdata/learn-more#Documentation) for more on what these layers are and what the flowline attributes entail.

# Bounding Boxes

First, a side note on bounding boxes. With the ongoing transition from the `sp` package to the `sf` package, there are a few stumbling blocks. Bounding boxes are one of them. As shown below, the `sf` bbox format is a named vector of class "bbox". The `sp` bbox format is a matrix with named dimensions. Many packages expect the sp format. the `ggmap` package expects yet another bbox format, much like `sf` but with different names. 

```{r bbox, message=FALSE}
library(sp)

sf_bbox <- st_bbox(basin)
sf_bbox
class(sf_bbox)

sp_bbox <- sp::bbox(sf::as_Spatial(basin))
sp_bbox
class(sp_bbox)

# Or without the sp::bbox
sp_bbox <- matrix(sf_bbox, 
                  byrow = FALSE, 
                  ncol = 2, 
                  dimnames = list(c("x", "y"), 
                                  c("min", "max")))
sp_bbox

ggmap_bbox <- setNames(sf_bbox, c("left", "bottom", "right", "top"))
ggmap_bbox
```

## Base R Plotting 

In order to maximize flexibility and make sure we understand what's going on with coordinate reference systems, the demonstration below shows how to use base R plotting with the package `prettymappr` and `rosm`. 

In this example, we have to plot just the geometry, extracted with `st_geometry` and we need to project the geometry into the plotting coordinate reference system, [EPSG:3857 also known as "web mercator"](https://en.wikipedia.org/wiki/Web_Mercator). The reason we have to make this transformation is that practically all basemap tiles are in this projection and reprojection of pre-rendered tiles doesn't look good. We do this with a simple `prep_layer` function.

The `prettymapr::prettymap()` function isn't strictly necessary, but it gives us nice margins, a scale bar, and a north arrow. The `rosm::osm.plot` and base `plot` commands put data onto the R plotting device so the first to be plotted is on the bottom. A couple hints here. `lwd` is line width. `pch` is point style. `cex` is an expansion factor. Colors shown below are [basic R colors.](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) the `rgb` function is handy for creating colors with transparency if that's of interest.

```{r plot}
prep_layer <- function(x) st_geometry(st_transform(x, 3857))

prettymapr::prettymap({
  rosm::osm.plot(sp_bbox, type = "stamenwatercolor", quiet = TRUE)
  
  plot(prep_layer(basin), 
       lwd = 2, add = TRUE)
  
  plot(prep_layer(flowline), 
       lwd = 1.5, col = "deepskyblue", add = TRUE)
  
  plot(prep_layer(dplyr::filter(flowline, streamorde > 2)), 
       lwd = 3, col = "darkblue", add = TRUE)
  
  us_nwis_layer <- prep_layer(upstream_nwis)
  
  plot(us_nwis_layer, 
       pch = 17, cex = 1.5, col = "yellow", add = TRUE)
  
  label_pos <- st_coordinates(us_nwis_layer)
  
  text(label_pos[,1],label_pos[,2], 
       upstream_nwis$identifier, 
       adj = c(-0.2, 0.5), cex = 0.7)
  
}, drawarrow = TRUE)
``` 

## Plotting with ggplot2

Below is a very similar example using [`ggmap`](https://github.com/dkahle/ggmap) and [`ggplot2` `geom_sf`](https://ggplot2.tidyverse.org/reference/ggsf.html). Note that ggmap takes case of projections for us, which should either make you happy because it _just works_ or very nervous because it _just works_. 

```{r ggmap}
library(ggmap)
library(ggplot2)

upstream_nwis[c("lon", "lat")] <- sf::st_coordinates(upstream_nwis)

basemap_toner <- get_map(source = "stamen", maptype = "toner", location = ggmap_bbox, zoom = 11)
basemap_terrain <- get_map(source = "stamen", maptype = "terrain-lines", location = ggmap_bbox, zoom = 11)
toner_map <- ggmap(basemap_toner)
terrain_map <- ggmap(basemap_terrain)

toner_map

terrain_map + geom_sf(data = basin,
                        inherit.aes = FALSE,
                        color = "black", fill = NA) + 
  geom_sf(data = flowline,
          inherit.aes = FALSE,
          color = "deepskyblue") +
  geom_sf(data = dplyr::filter(flowline, streamorde > 2),
          inherit.aes = FALSE,
          color = "darkblue") +
  geom_sf(data = upstream_nwis, inherit.aes = FALSE, color = "red") + 
  geom_text(data = upstream_nwis, aes(label = identifier, x = lon, y = lat),
            hjust = 0, size=2.5, nudge_x = 0.02, col = "black")
```

Hopefully these examples give a good head start to plotting NHDPlus data. Please submit questions via github issues for more!! Pull requests on this vignette are more than welcome if you have additions or suggestions.

```{r teardown, include=FALSE}
options(oldoption)
```
