---
title: "3DHP California Demo"
format: 
  html:
    self-contained: false
    embed-resources: false
execute:
  cache: true
---

```{r}
#| echo: false
library(widgetframe)
```

## 3DHP is at the heart of water data on the Web

Start from a mainstem river: [https://explorer.geoconnex.us/mainstems/1805986](https://explorer.geoconnex.us/mainstems/1805986) or [https://geoconnex.us/ref/mainstems/1805986 
](https://geoconnex.us/ref/mainstems/1805986) 

First just grab the mainstem from 3DHP.

```{r}

mainstem <- "https://geoconnex.us/ref/mainstems/1805986"

stan <- nhdplusTools::get_3dhp(ids = mainstem, type = "flowline")

names(stan)

frameWidget(mapview::mapview(stan)@map)

```

Mainstems let us do some other neat stuff!

Note that we can get IDs from other datasets that are associated with 3DHP features with reference mainstems.

```{r}

ms <- sf::read_sf(mainstem)

ms

# this gives us a feature with geometry and a comid attribute!
outlet <- sf::read_sf(ms$outlet_nhdpv2_comid)

frameWidget(mapview::mapview(outlet)@map)

```

Now we'll get a basin boundary... from NHDPlusV2 but it's a useful tool here.

```{r}
basin <- nhdplusTools::get_nldi_basin(nldi_feature = list(featureSource = "comid", featureId = outlet$comid))

frameWidget(mapview::mapview(basin)@map)

```

Now let's get all the 3DHP data for the basin.

```{r}

full_stan <- nhdplusTools::get_3dhp(sf::st_buffer(basin, units::set_units(1, "km")), type = "flowline")

nrow(full_stan)

frameWidget(mapview::mapview(full_stan)@map)

```

## 3DHP data are available "services first"!

Now let's move to a basin where there is 3DHP data available.

```{r}

mainstem <- "https://geoconnex.us/ref/mainstems/328091"

hols <- nhdplusTools::get_3dhp(ids = mainstem, type = "flowline")

names(hols)

sum(is.na(hols$catchmentid3dhp))
sum(!is.na(hols$catchmentid3dhp))

frameWidget(mapview::mapview(hols)@map)

```

We can get catchments for some of these features!

```{r}

cats <- hols$catchmentid3dhp[!is.na(hols$catchmentid3dhp)]

hols_cats <- nhdplusTools::get_3dhp(ids = cats, type = "catchment")

frameWidget(mapview::mapview(hols_cats)@map)

```

Or use an area of interest to get a bunch of catchments.

```{r}

aoi <- hols_cats[20,] |>
    sf::st_bbox() |>
    sf::st_as_sfc() |>
    nhdplusTools::get_huc(type = "huc10")

frameWidget(mapview::mapview(aoi)@map)

```

```{r}

all_holston <- nhdplusTools::get_3dhp(sf::st_buffer(aoi, units::set_units(1, "km")), type = "flowline")

all_holston <- all_holston[is.na(all_holston$workunitid) | all_holston$workunitid != "NHD", ]

frameWidget(mapview::mapview(all_holston)@map)

par(mar = c(0,0,0,0))
plot(sf::st_geometry(all_holston), lwd = all_holston$streamorder / 2)

```

## 3DHP data allow advanced analyses out of the box! 

```{r}

boundary <- nhdplusTools::get_huc(id = "02050302", type = "huc08")

juni <- nhdplusTools::get_3dhp(sf::st_buffer(boundary, units::set_units(1, "km")), type = "flowline")

main_unit <- names(which.max(table(juni$workunitid)))

juni <- juni[!is.na(juni$workunitid) & juni$workunitid == main_unit,]

nrow(juni)

frameWidget(mapview::mapview(list(boundary, juni))@map)

```

Now we'll select out some of the basic network attributes and do a navigation:

NOTE: A non-dendritic network is available in annual static snapshots and plans exist to incorporate that network information into web services in a future release.

```{r}

names(juni)

juni_sub <- dplyr::select(juni, id = hydrosequence, toid = dnhydrosequence, orig_id3dhp = id3dhp, mainstemid, lengthkm, catchmentid3dhp, arbolatesum)

hydroloom::check_hy_graph(juni_sub)

main_outlet <- juni_sub[!is.na(juni_sub$arbolatesum) & juni_sub$arbolatesum == max(juni_sub$arbolatesum, na.rm = TRUE),]

nrow(main_outlet)

reachable <- hydroloom::navigate_network_dfs(juni_sub, main_outlet$id, direction = "up")

length(unlist(reachable))

juni_reachable <- juni_sub[juni_sub$id %in% unlist(reachable),]

plot(sf::st_geometry(juni_reachable), lwd = (20 * juni_reachable$arbolatesum / (max(juni_reachable$arbolatesum))))

```

Now let's circle back to why 3DHP being core to the IoW is useful with an example of stream gages.

```{r}

outlet_mainstem <- juni_reachable[juni_reachable$mainstemid == main_outlet$mainstemid,]

gages <- nhdplusTools::get_geoconnex_reference(boundary, type = "gage")

# only gages on the mainstem we have
gages <- gages[!is.na(gages$mainstem_uri) & gages$mainstem_uri == outlet_mainstem$mainstemid[1],]

# get the actual flowline each gage is on in this version of 3DHP
indexes <- hydroloom::index_points_to_lines(dplyr::select(outlet_mainstem, id3dhp = orig_id3dhp), gages)
gages$point_id <- seq_len(nrow(gages))
gages <- dplyr::left_join(gages, indexes, by = "point_id")

frameWidget(mapview::mapview(list(outlet_mainstem, gages))@map)

```